---
title: "CodeBook"
author: "Karthik"
date: "2022-08-18"
output: html_document
---

```{r setup, include=FALSE}
library(dplyr)
library(tidyr)
source("~/Downloads/UCI HAR Dataset 2/run_analysis.R")
```

# Getting and cleaning data assignment 1

For this assignment we are given data from an experiment in which 30 subjects were asked to wear a smart phone and perform six different activities. The signals from their smart phone's accelerometer and gyroscope were recorded in three dimensions. The signals were then applied through a noise filter and transformed into other variables. 

The goal of the assignment was merge the training and test datasets, extract only the mean and std deviation for the measurements and create a new independent tidy dataset

The below process is identical for the testing and training datasets until they are merged

### Getting subject ids  

```{r subject_ids}
test_subject_ids <- read.table("test/subject_test.txt")
colnames(test_subject_ids) <- c("subject_id")
unique(test_subject_ids)
```

### Getting activity names  

```{r activity_labels}
activity_names <- read.table("activity_labels.txt")
activity_names
```

### Getting activity ids 
```{r activity_ids}
test_activity <- read.table("test/y_test.txt")
head(test_activity)
```

### Merge activity_id with activity_name
When reading a text file R automatically creates a column name if one doesn't exist like V1, V2 etc. I am renaming them to activity_name and activity_id, so it's easy to remember and reference them 

``` {r activity_merge}
colnames(test_activity) <- c("activity_id")
colnames(activity_names) <- c("activity_id", "activity_name")
test_activity_names <- merge(test_activity, activity_names)$activity_name
head(test_activity_names)
```

test_activity_names now has all the activity names instead of activity_ids

### Getting the measurement names 
``` {r measurement_name}
variable_names <- read.table("features.txt")
head(variable_names)
variable_names[106,]
```

The variable names above has parenthesis and comma's in it. I used the gsub method to remove the commas and parenthesis. The measurements that start with "t" and "f" I replaced with Time and Frequency respectively. For Mag replaced it with Magnitude.  

``` {r cleaning_var_names}
variable_names$V2 <- gsub(",", "-", variable_names$V2)
variable_names$V2 <- gsub("\\(", "", variable_names$V2)
variable_names$V2 <- gsub("\\)", "", variable_names$V2)
variable_names$V2 <- gsub("^t", "Time", variable_names$V2)
variable_names$V2 <- gsub("^f", "Frequency", variable_names$V2)
variable_names$V2 <- gsub("^Mag", "Magnitude", variable_names$V2)
variable_names[106,]
```

### Reading the test signals 

``` {r reading_signals}
test_signals <- read.table("test/X_test.txt")
test_signals[1:5,1:5]
```

The column names are numbered from V1 to V561 and make it really hard for later analysis. I fixed the column names for the signal data using the variable_names that I cleaned up above

``` {r fixing_col_names}
colnames(test_signals) <- variable_names$V2
head(names(test_signals))
```

### Making the test data frame 

``` {r test_df}
test_data <- data.frame(test_subject_ids, test_activity_names, test_signals)
test_tbl <- tbl_df(test_data)
test_tbl
```

I used the similar approach to put together the training data 

### Merging the test and the training data sets 

The next step was to merge the two data sets. So I put together small helper functions that did the work described above into run_analysis.R. The readData function takes an input parameter either "test" or "train" and then loads the data for that into a data frame by appropriately naming the variables and activity labels.

The two data frames can be merged using the rbind function 

``` {r merging} 
train <- readData("test")
test <- readData("train")
merged <- rbind(test, train)
```

### Filtering the required rows

``` {r filtering}
merged_tbl <- tbl_df(merged)
merged_tbl %>% select(subject_id, activity_label, matches("mean|std")) %>% group_by(subject_id, activity_label) %>% summarize_all(.funs = c(mean = "mean"))
```


### Writing it out
In this class I learned that csv are much easier to work with compared with other formats. Given that we have a choice now to clean up this data, I used the write.csv method to write the data out. 
``` {r writing_it_out}
write.csv(merged_tbl, "output.csv")
```